Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2018-03-19 poster/qnmlPoster.tex
more poster

\mode<presentation>
{
  \usetheme{hiitposter}
}
%\graphicspath{{imgs/}}
\usepackage{amsmath,amsthm, amssymb,mathrsfs}
\usepackage{color}
\usepackage{times}
\usepackage{subfigure}
\usepackage{bm}
\usepackage{url}
\usepackage[numbers]{natbib}
\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}        
%\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
%\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
%\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}

@@

\begin{block}{Background: Structure Learning of Bayesian Networks}
\begin{columns}[T]
   \begin{column}{0.3\textwidth} % first column {{{
     \heading{Bayesian Networks}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
     %\item Here, each $X_i$ is a categorical random variable.
     \item Consist of:
     \begin{itemize}
     \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
     \end{itemize}
     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid G_i,\theta_i)$
     \end{itemize}     
   \end{column}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){450}
   \end{column} % }}}
   \begin{column}{0.3\textwidth}% third column {{{
     \heading{Scoring Functions}
     \begin{itemize}

	 \vspace*{12pt}
	 \setlength\itemsep{1em}     
     \item[BDeu] Bayesian marginal likelihood based on Dirichlet priors. Depends on a single 
     hyperparameter $\alpha >0$ called equivalent sample size. 
     \item[BIC] Maximized likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
     \item[fNML] Factorized Normalized Maximum likelihood. Maximized likelihood with penalty defined via \textit{regret} functions.
     \end{itemize}

@@

   \begin{column}{0.3\textwidth} % second column {{{
	\heading{qNML: Definition}
	\vspace*{12pt}
	\begin{itemize}
	\item We would like to choose a model $G$ maximizing	
	\begin{equation}
	P_{NML}(D;G)=\frac{P(D \mid \hat\theta , G)}{\sum_{D'}{P(D' \mid \hat  \theta,G)}},
	\end{equation}where $\hat\theta$ denotes the maximum likelihood parameters.
	\end{itemize}

@@

      \heading{qNML: Properties}

   \end{column} % end of first column }}}
   % \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
   %    \color{myPurple} \line(0,1){600}
   % \end{column} % }}}
   % \begin{column}{0.3\textwidth} % third column {{{
   % \end{column} % end of third column }}}

@@

    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){700}
   \end{column} % }}}

@@

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){700}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Conclusions}

>>>>

\mode<presentation>
{
  \usetheme{hiitposter}
}
%\graphicspath{{imgs/}}
\usepackage{amsmath,amsthm, amssymb,mathrsfs}
\usepackage{color}
\usepackage{times}
\usepackage{subfigure}
\usepackage{bm}
\usepackage{url}
\usepackage[numbers]{natbib}
\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
\usepackage{blkarray}        
%\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
%\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
%\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}

@@

\begin{block}{Background: Structure Learning of Bayesian Networks}
\begin{columns}[T]
   \begin{column}{0.3\textwidth} % first column {{{
     \heading{Bayesian Networks}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
     %\item Here, each $X_i$ is a categorical random variable.
     \item Consist of:
     \begin{itemize}
     \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
     \end{itemize}
     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid X_{G_i},\theta_i)$
     \end{itemize}     
   \end{column}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){450}
   \end{column} % }}}
   \begin{column}{0.3\textwidth}% third column {{{
     \heading{Scoring Functions}
     \begin{itemize}
	 \vspace*{12pt}
	 \setlength\itemsep{1em}     
     \item[BDeu] Bayesian marginal likelihood based on Dirichlet priors. Depends on a single 
     hyperparameter $\alpha >0$ called equivalent sample size. 
     \item[BIC] Maximized log-likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
     \item[fNML] Factorized Normalized Maximum likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
     \end{itemize}

@@

   \begin{column}{0.3\textwidth} % second column {{{
	\heading{qNML: Idea}
	\vspace*{12pt}
	\begin{itemize}
	\item Following the \textit{MDL-principle}, we would like to pick a model $G$ maximizing

	\begin{equation}
	\log P_{NML}(D;G) = \underbrace{\log P(D \mid \hat\theta , G)}_{\text{Maximized log-likelihood}} - \underbrace{\log \sum_{D'}P(D' \mid \hat  \theta,G)}_{:= \text{ regret. } \textbf{Intractable!}}, \notag
	\end{equation}%where $\hat\theta$ denotes the maximum likelihood parameters.

	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
	%\item Justification from \textit{Minimum description length} principle.
	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
	$$
	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
	$$
	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
	\vspace*{12pt}
	\end{itemize}
	\begin{minipage}{0.45\textwidth}
	\begin{align*}
	\begin{blockarray}{c|ccc}
	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
	\begin{block}{(c|ccc)}
  	1 & 1 & 1 & 1  \\
  	0 & 1 & 0 & 0 \\
  	1 & 1 &0 & 0  \\
  	1 & 1 & 1 & 1  \\
	\end{block}
	\end{blockarray} 
	\ \to \ 
	\begin{blockarray}{c}
	\phantom{...}\\
	\begin{block}{(c)}
  	0   \\
  	1  \\
  	2  \\
  	0   \\
	\end{block}
	\end{blockarray}
 	\end{align*}
	\end{minipage}\quad
	\begin{minipage}{0.45\textwidth}
	\fbox{\parbox{\textwidth}{
	\begin{itemize}
	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
	\end{itemize}}}  
	\end{minipage}

@@

   	\heading{qNML: Definition \& Properties}
   	Definition of qNML score:
    	\begin{empheq}[box={\mybluebox[4pt][4pt]}]{equation}	    		
	s^{qNML} := \sum_{i = 1}^d \log \frac{P^1_{NML}(D_{i,G_i})}{P^1_{NML}(D_{G_i})},\notag
	\end{empheq}$\log P^1_{NML}(D) = \log P(D \mid \hat{\theta}) - reg(n,r).$

	\begin{itemize}
	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
	\begin{align*}
    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
	\end{align*}
where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
	\end{itemize}
	\vspace*{12pt}
	qNML score is:
	\fbox{\parbox{0.95\textwidth}{
	\begin{itemize}	
	\item \textcolor{teal}{\textbf{hyper-parameter free}}
	\item \textcolor{teal}{\textbf{consistent}}
	\item \textcolor{teal}{\textbf{score equivalent}}
	\item \textcolor{teal}{\textbf{regular}}
	\item qNML agrees with NML for every network composed of fully connected sub-graphs.
	\end{itemize}}}

@@

	  \begin{minipage}{0.45\textwidth}
	  \begin{figure}[h]

		\includegraphics[width=\columnwidth]{shd_all.pdf}
		\caption{SHD as a function of sample size.}
		\label{fig:all_shd}
		\end{figure}
	  \end{minipage}
	  \begin{minipage}{0.45\textwidth}
	  \begin{itemize}
	  \item Data generated from real world DAG structures.
	  \item Exact DP structure learning algorithm.
	  \end{itemize}
	  \begin{figure}[h]
		\includegraphics[width=\columnwidth]{shd_rank_all.pdf}
		\caption{Average ranks for each score.}
		\label{fig:all_shd}
		\end{figure}
	  \end{minipage}

	\vspace*{12pt}
	\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){600}
   \end{column} % }}}

@@

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){600}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Conclusions}
